---
title: "Getting Started with gretaR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with gretaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = torch::torch_is_installed()
)
```

## Overview

**gretaR** is a Bayesian statistical modelling package for R built on the
`torch` backend. It provides an intuitive, R-native syntax for defining
probabilistic models, which are then compiled to differentiable torch
computations and sampled using Hamiltonian Monte Carlo (HMC) or the
No-U-Turn Sampler (NUTS).

**Key features:**

- Pure R — no Python or reticulate dependency
- GPU-accelerated inference via torch
- Familiar R syntax: standard operators, functions, and assignment
- Output compatible with `posterior`, `bayesplot`, and `loo`

## Installation

```{r install, eval = FALSE}
# Install from GitHub (development version)
# remotes::install_github("maxmoldovan/gretaR")

# Ensure torch is installed
torch::install_torch()
```

## Example 1: Estimating a Normal Mean

Suppose we observe data from a normal distribution and want to estimate
the mean and standard deviation.

```{r normal-model}
library(gretaR)

# Simulate data
set.seed(42)
y_obs <- rnorm(100, mean = 3, sd = 1.5)

# Define priors
mu <- normal(0, 10)
sigma <- half_cauchy(5)

# Wrap observed data
y <- as_data(y_obs)

# Define likelihood
distribution(y) <- normal(mu, sigma)

# Compile model
m <- model(mu, sigma)
print(m)
```

```{r normal-sample, eval = FALSE}
# Draw posterior samples
draws <- mcmc(m, n_samples = 1000, warmup = 1000, chains = 4)

# Summarise
summary(draws)
```

## Example 2: Bayesian Linear Regression

```{r linreg-model}
# Simulate data
set.seed(123)
n <- 100
x_obs <- rnorm(n)
y_obs <- 2 + 3 * x_obs + rnorm(n, 0, 0.5)

# Reset model state for new model
reset_gretaR_env()

# Priors
alpha <- normal(0, 10)
beta <- normal(0, 10)
sigma <- half_cauchy(2)

# Data
x <- as_data(x_obs)
y <- as_data(y_obs)

# Linear predictor
mu <- alpha + beta * x

# Likelihood
distribution(y) <- normal(mu, sigma)

# Compile
m <- model(alpha, beta, sigma)
print(m)
```

```{r linreg-sample, eval = FALSE}
# Sample
draws <- mcmc(m, n_samples = 1000, warmup = 1000, chains = 4)
summary(draws)

# Visualise traces (requires bayesplot)
# plot(draws, type = "trace")
```

## Workflow Summary

1. **Wrap data** with `as_data()`
2. **Define priors** using distribution functions (`normal()`, `half_cauchy()`, etc.)
3. **Build the model** using standard R operations (`+`, `*`, `log()`, etc.)
4. **Assign the likelihood** with `distribution(y) <- ...`
5. **Compile** with `model()`
6. **Sample** with `mcmc()`
7. **Analyse** using `posterior`, `bayesplot`, or `loo`

## Available Distributions

| Function | Distribution | Support |
|---|---|---|
| `normal(mean, sd)` | Normal | $(-\infty, \infty)$ |
| `half_normal(sd)` | Half-Normal | $(0, \infty)$ |
| `half_cauchy(scale)` | Half-Cauchy | $(0, \infty)$ |
| `student_t(df, mu, sigma)` | Student-t | $(-\infty, \infty)$ |
| `uniform(lower, upper)` | Uniform | $(a, b)$ |
| `exponential(rate)` | Exponential | $(0, \infty)$ |
| `gamma_dist(shape, rate)` | Gamma | $(0, \infty)$ |
| `beta_dist(alpha, beta)` | Beta | $(0, 1)$ |
| `bernoulli(prob)` | Bernoulli | $\{0, 1\}$ |
| `binomial(size, prob)` | Binomial | $\{0, \ldots, n\}$ |
| `poisson_dist(rate)` | Poisson | $\{0, 1, 2, \ldots\}$ |
| `multivariate_normal(mean, cov)` | Multivariate Normal | $\mathbb{R}^k$ |

## Choosing a Sampler

- **NUTS** (default): Adaptive, no tuning of leapfrog steps needed. Recommended
  for most models.
- **HMC**: Fixed number of leapfrog steps. Useful for debugging or when NUTS
  tree depth is a concern.

```{r sampler-choice, eval = FALSE}
# NUTS (default)
draws <- mcmc(m, sampler = "nuts")

# HMC with 25 leapfrog steps
draws <- mcmc(m, sampler = "hmc", n_leapfrog = 25)
```

## Next Steps

- Explore hierarchical models (coming in v0.2)
- Use `bayesplot::mcmc_trace()` and `bayesplot::mcmc_dens_overlay()` for
  diagnostics
- Compare models with `loo::loo()` on the posterior draws
