---
title: "Getting Started with gretaR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with gretaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = torch::torch_is_installed()
)
```

## Overview

**gretaR** is a Bayesian statistical modelling package for R built on the
`torch` backend. It provides an intuitive, R-native syntax for defining
probabilistic models, which are then compiled to differentiable torch
computations and sampled using Hamiltonian Monte Carlo (HMC) or the
No-U-Turn Sampler (NUTS).

**Key features:**

- Pure R -- no Python or reticulate dependency
- Native torch (libtorch) backend -- CPU today, GPU/MPS on the roadmap
- Familiar R syntax: standard operators, functions, and assignment
- Output compatible with `posterior`, `bayesplot`, and `loo`

## Installation

```{r install, eval = FALSE}
# Install from GitHub (development version)
# remotes::install_github("max578/gretaR")

# Ensure torch is installed
torch::install_torch()
```

## Example 1: Estimating a Normal Mean

Suppose we observe data from a normal distribution and want to estimate
the mean and standard deviation.

```{r normal-model}
library(gretaR)

# Simulate data
set.seed(42)
y_obs <- rnorm(100, mean = 3, sd = 1.5)

# Define priors
mu <- normal(0, 10)
sigma <- half_cauchy(5)

# Wrap observed data
y <- as_data(y_obs)

# Define likelihood
distribution(y) <- normal(mu, sigma)

# Compile model
m <- model(mu, sigma)
print(m)
```

```{r normal-sample}
# Draw posterior samples
fit1 <- mcmc(m, n_samples = 500, warmup = 500, chains = 2)

# Summarise
summary(fit1)
```

```{r normal-trace, fig.cap = "Trace plot for the normal-mean model. Both chains overlap and show no drift or stuck regions across the sampled range, the visual signature of a converged NUTS run that agrees with the printed R-hat above."}
plot(fit1, type = "trace")
```

The two chains mix over the same range of `mu` and `sigma` from the first
sampled draw, with no visible trend or one chain sitting apart from the
other -- consistent with the R-hat close to 1 and the effective sample
size reported by `summary(fit1)`.

## Example 2: Bayesian Linear Regression

```{r linreg-model}
# Simulate data
set.seed(123)
n <- 100
x_obs <- rnorm(n)
y_obs <- 2 + 3 * x_obs + rnorm(n, 0, 0.5)

# Reset model state for new model
reset_gretaR_env()

# Priors
alpha <- normal(0, 10)
beta <- normal(0, 10)
sigma <- half_cauchy(2)

# Data
x <- as_data(x_obs)
y <- as_data(y_obs)

# Linear predictor
mu <- alpha + beta * x

# Likelihood
distribution(y) <- normal(mu, sigma)

# Compile
m <- model(alpha, beta, sigma)
print(m)
```

```{r linreg-sample}
# Sample
fit2 <- mcmc(m, n_samples = 500, warmup = 500, chains = 2)
summary(fit2)
```

```{r linreg-density, fig.cap = "Posterior density overlays for the regression intercept, slope and residual scale. Each chain's density sits on top of the others and each parameter's mass is centred close to its simulated true value (2, 3 and 0.5 respectively)."}
plot(fit2, type = "density")
```

The three densities are unimodal, chain-to-chain agreement is close, and
the posterior mass for `alpha`, `beta` and `sigma` brackets the values used
to simulate `y_obs`, confirming the model recovers the generating
parameters on this synthetic dataset.

## Workflow Summary

1. **Wrap data** with `as_data()`
2. **Define priors** using distribution functions (`normal()`, `half_cauchy()`, etc.)
3. **Build the model** using standard R operations (`+`, `*`, `log()`, etc.)
4. **Assign the likelihood** with `distribution(y) <- ...`
5. **Compile** with `model()`
6. **Sample** with `mcmc()`
7. **Analyse** using `posterior`, `bayesplot`, or `loo`

## Available Distributions

| Function | Distribution | Support |
|---|---|---|
| `normal(mean, sd)` | Normal | $(-\infty, \infty)$ |
| `half_normal(sd)` | Half-Normal | $(0, \infty)$ |
| `half_cauchy(scale)` | Half-Cauchy | $(0, \infty)$ |
| `student_t(df, mu, sigma)` | Student-t | $(-\infty, \infty)$ |
| `uniform(lower, upper)` | Uniform | $(a, b)$ |
| `exponential(rate)` | Exponential | $(0, \infty)$ |
| `gamma_dist(shape, rate)` | Gamma | $(0, \infty)$ |
| `beta_dist(alpha, beta)` | Beta | $(0, 1)$ |
| `bernoulli(prob)` | Bernoulli | $\{0, 1\}$ |
| `binomial_dist(size, prob)` | Binomial | $\{0, \ldots, n\}$ |
| `poisson_dist(rate)` | Poisson | $\{0, 1, 2, \ldots\}$ |
| `multivariate_normal(mean, cov)` | Multivariate Normal | $\mathbb{R}^k$ |

## Choosing a Sampler

- **NUTS** (default): Adaptive, no tuning of leapfrog steps needed. Recommended
  for most models.
- **HMC**: Fixed number of leapfrog steps. Useful for debugging or when NUTS
  tree depth is a concern.

```{r sampler-choice, eval = FALSE}
# NUTS (default)
draws <- mcmc(m, sampler = "nuts")

# HMC with 25 leapfrog steps
draws <- mcmc(m, sampler = "hmc", n_leapfrog = 25)
```

## Next Steps

- Explore hierarchical models in `vignette("hierarchical-models", package = "gretaR")`
- Use `bayesplot::mcmc_trace()` and `bayesplot::mcmc_dens_overlay()` for
  diagnostics
- Compare models with `loo::loo()` on the posterior draws
