---
title: "Hierarchical Models with gretaR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical Models with gretaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = torch::torch_is_installed()
)
```

## Overview

Hierarchical (multi-level) models are a key strength of Bayesian inference.
gretaR supports hierarchical models through the `[` indexing operator, which
lets you define group-level parameters and index them by observation.

## Random Intercepts Model

Consider data from `J` groups, each with `n_j` observations:

$$y_{ij} = \alpha_j + \epsilon_{ij}, \quad \epsilon_{ij} \sim N(0, \sigma^2)$$

$$\alpha_j \sim N(\mu, \tau^2)$$

```{r random-intercepts}
library(gretaR)

# Simulate grouped data
set.seed(42)
n_groups <- 5
n_per_group <- 20
n <- n_groups * n_per_group
group_id <- rep(1:n_groups, each = n_per_group)

true_mu <- 5
true_tau <- 2
true_sigma <- 1
true_alpha <- rnorm(n_groups, true_mu, true_tau)
y_obs <- rnorm(n, true_alpha[group_id], true_sigma)

# Define gretaR model
mu <- normal(0, 10)
tau <- half_cauchy(5)
alpha <- normal(mu, tau, dim = c(n_groups, 1))
sigma <- half_cauchy(5)

y <- as_data(y_obs)
fitted_vals <- alpha[group_id]  # Index into group-level parameters
distribution(y) <- normal(fitted_vals, sigma)

m <- model(mu, tau, sigma, alpha)
print(m)
```

The `alpha[group_id]` syntax creates an operation node in the DAG that
selects the appropriate group-level parameter for each observation.
This is equivalent to the indexing notation used in Stan and JAGS.

```{r fit-random-intercepts, eval = FALSE}
# Fit with NUTS
draws <- mcmc(m, n_samples = 1000, warmup = 1000, chains = 4)
summary(draws)
```

## Random Intercepts and Slopes

Extend to include a group-level slope:

$$y_{ij} = \alpha_j + \beta_j x_{ij} + \epsilon_{ij}$$

$$\alpha_j \sim N(\mu_\alpha, \tau_\alpha^2), \quad \beta_j \sim N(\mu_\beta, \tau_\beta^2)$$

```{r random-slopes}
reset_gretaR_env()
set.seed(123)

n_groups <- 8
n_per_group <- 15
n <- n_groups * n_per_group
group_id <- rep(1:n_groups, each = n_per_group)
x_obs <- rnorm(n)

true_mu_a <- 3; true_tau_a <- 1.5
true_mu_b <- 2; true_tau_b <- 0.5
true_sigma <- 0.8

true_a <- rnorm(n_groups, true_mu_a, true_tau_a)
true_b <- rnorm(n_groups, true_mu_b, true_tau_b)
y_obs <- true_a[group_id] + true_b[group_id] * x_obs + rnorm(n, 0, true_sigma)

# Model
mu_a <- normal(0, 10)
mu_b <- normal(0, 10)
tau_a <- half_cauchy(5)
tau_b <- half_cauchy(5)
alpha <- normal(mu_a, tau_a, dim = c(n_groups, 1))
beta <- normal(mu_b, tau_b, dim = c(n_groups, 1))
sigma <- half_cauchy(5)

x <- as_data(x_obs)
y <- as_data(y_obs)

mu_y <- alpha[group_id] + beta[group_id] * x
distribution(y) <- normal(mu_y, sigma)

m <- model(mu_a, mu_b, tau_a, tau_b, sigma, alpha, beta)
print(m)
```

## Using MAP for Quick Estimates

For quick model checking, use `opt()` to find the MAP estimate:

```{r map-estimate, eval = FALSE}
fit <- opt(m, verbose = TRUE)
fit$par
```

## Using Variational Inference

For faster approximate inference:

```{r vi-estimate, eval = FALSE}
fit <- variational(m, method = "meanfield", max_iter = 3000)
fit$mean
fit$sd
summary(fit$draws)
```

## Partially-Centred Random Effects with `random_effect()`

The centred and non-centred parameterisations above are the two ends of a
continuum: `random_effect()` builds the same group-level prior,
$u_j \sim N(\mathrm{mean}, \mathrm{sd}^2)$, through a `centring` weight
$w \in [0, 1]$ that interpolates the sampling geometry (`0` non-centred,
`1` centred, anything between is a partial reparameterisation) without
changing what is being estimated. It returns a `$latent` block, which must
be passed to `model()`, and a `$effect`, the per-observation value that
enters the linear predictor.

```{r random-effect-verb}
reset_gretaR_env()
set.seed(42)

n_groups <- 5
n_per_group <- 20
n <- n_groups * n_per_group
group_id <- rep(1:n_groups, each = n_per_group)

true_mu <- 5
true_tau <- 2
true_sigma <- 1
true_alpha <- rnorm(n_groups, true_mu, true_tau)
y_obs <- rnorm(n, true_alpha[group_id], true_sigma)

mu <- normal(0, 10)
tau <- half_cauchy(5)
sigma <- half_cauchy(5)

# A centring weight of 0.5 is a reasonable starting point when groups have
# a moderate amount of data each, as here (n_per_group = 20).
re <- random_effect(group_id, n_groups, sd = tau, centring = 0.5, mean = mu)

y <- as_data(y_obs)
distribution(y) <- normal(re$effect, sigma)

m <- model(mu, tau, sigma, re$latent)
print(m)
```

```{r random-effect-sample}
fit_re <- mcmc(m, n_samples = 300, warmup = 300, chains = 2)
summary(fit_re)
```

```{r random-effect-trace, fig.cap = "Trace plot for the partially-centred random-effect model (centring = 0.5). Both chains cover the same range for every group-level and hyper-parameter with no persistent separation, the pattern expected once the centring weight is matched to how much data each group carries."}
plot(fit_re, type = "trace")
```

The hyper-parameters `mu` and `tau` and all five latent group terms mix
without the funnel-shaped stickiness a badly-matched centring weight
produces; `centring` is a construction-time choice here, not something
`random_effect()` tunes automatically -- a group with very few
observations relative to `tau` mixes better closer to `w = 0`, and a
richly-informed group closer to `w = 1`.

## Formula Interface for Simple Models

For standard GLMs, use the formula interface:

```{r formula-glm, eval = FALSE}
# Simple linear regression via formula
dat <- data.frame(y = y_obs, x = x_obs, group = factor(group_id))
fit <- gretaR_glm(y ~ x, data = dat, family = "gaussian", sampler = "map")
print(fit)
```

## Tips for Hierarchical Models

1. **Parameterisation**: Use `random_effect()` with a low `centring`
   weight (closer to non-centred) for better sampling when data are sparse
   within groups, and a higher weight (closer to centred) when groups are
   richly informed; hand-rolling the non-centred form directly is
   equivalent to `centring = 0` but loses the ability to move along the
   continuum.
2. **Priors on variance components**: `half_cauchy()` is a sensible
   default for group-level standard deviations (Gelman 2006).
3. **Convergence**: Check R-hat and ESS. Hierarchical models may need
   longer warmup periods.
4. **Quick checks**: Use `opt()` or `variational()` to verify the model
   before running full MCMC.

## References

Gelman, A. (2006). Prior distributions for variance parameters in
hierarchical models (comment on article by Browne and Draper). *Bayesian
Analysis*, 1(3), 515--534. `doi:10.1214/06-BA117A`
