---
title: "Migrating from greta to gretaR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from greta to gretaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

**gretaR** is a ground-up reimplementation of the
[greta](https://greta-stats.org/) probabilistic programming interface for R.
It preserves the elegant, intuitive API that made greta popular while removing
its most significant pain point — the Python/TensorFlow dependency — and adding
modern inference capabilities.

Three reasons to migrate:

1. **No Python dependency.** gretaR uses
   [torch for R](https://torch.mlverse.org/) as its computation backend.
   Installation is a single `install.packages("gretaR")` call with no
   `reticulate`, no conda environments, and no TensorFlow version conflicts.

2. **Broader inference methods.** In addition to HMC/NUTS, gretaR ships MAP
   optimisation (`opt()`), Laplace approximation (`laplace()`), and automatic
   differentiation variational inference (`vi()`). These are useful for rapid
   prototyping, model comparison, and large-data settings where full MCMC is
   impractical.

3. **Active, focused development.** gretaR is under active development with a
   clear roadmap covering sparse matrix support, formula-based model
   specification, and extended distribution families.

## Side-by-side API comparison

The table below maps the most common greta operations to their gretaR
equivalents. Where the syntax is identical, migration requires no code changes
at all.

| Task | greta | gretaR | Notes |
|:-----|:------|:-------|:------|
| Data wrapping | `greta::as_data(x)` | `gretaR::as_data(x)` | Identical API |
| Variable creation | `greta::variable(lower, upper, dim)` | `gretaR::variable(lower, upper, dim)` | Identical API |
| Normal distribution | `greta::normal(mean, sd)` | `gretaR::normal(mean, sd)` | Identical API |
| Half-Cauchy | `greta::cauchy(0, scale, truncation = c(0, Inf))` | `gretaR::half_cauchy(scale)` | Dedicated constructor in gretaR |
| Student-t | `greta::student(df, mu, sigma)` | `gretaR::student_t(df, mu, sigma)` | Name differs slightly |
| Gamma | `greta::gamma(shape, rate)` | `gretaR::gamma_dist(shape, rate)` | Renamed to avoid masking `base::gamma()` |
| Beta | `greta::beta(shape1, shape2)` | `gretaR::beta_dist(alpha, beta)` | Renamed to avoid masking `base::beta()` |
| Poisson | `greta::poisson(lambda)` | `gretaR::poisson_dist(rate)` | Renamed to avoid masking `stats::poisson()` |
| Uniform | `greta::uniform(min, max)` | `gretaR::uniform(lower, upper)` | Parameter names differ |
| Bernoulli | `greta::bernoulli(prob)` | `gretaR::bernoulli(prob)` | Identical API |
| Binomial | `greta::binomial(size, prob)` | `gretaR::binomial(size, prob)` | Identical API |
| Exponential | `greta::exponential(rate)` | `gretaR::exponential(rate)` | Identical API |
| Multivariate normal | `greta::multivariate_normal(mean, Sigma)` | `gretaR::multivariate_normal(mean, covariance)` | Identical API |
| Dirichlet | `greta::dirichlet(alpha)` | `gretaR::dirichlet(concentration)` | Parameter name differs |
| LKJ correlation | `greta::lkj_correlation(eta, dimension)` | `gretaR::lkj_correlation(eta, dim)` | Parameter name differs |
| Distribution assignment | `distribution(y) <- normal(mu, sigma)` | `distribution(y) <- normal(mu, sigma)` | Identical syntax |
| Model creation | `greta::model(...)` | `gretaR::model(...)` | Identical API |
| MCMC sampling | `greta::mcmc(model)` | `gretaR::mcmc(model)` | Arguments differ (see below) |

### Distributions unique to gretaR

gretaR provides several additional distribution constructors not available in
greta:

- `half_normal(sd)` — half-normal distribution on the positive reals
- `half_cauchy(scale)` — first-class constructor (no truncation workaround)
- `negative_binomial(size, prob)` — negative binomial for overdispersed counts
- `lognormal(meanlog, sdlog)` — log-normal distribution
- `cauchy(location, scale)` — full (two-sided) Cauchy distribution
- `wishart(df, scale_matrix)` — Wishart distribution for covariance matrices

## Key differences

### 1. No Python dependency

greta requires a working Python installation with TensorFlow and TensorFlow
Probability. This is the single most common source of installation failures.
gretaR replaces the entire Python stack with
[torch for R](https://torch.mlverse.org/), which is a native R package with
pre-built binaries for macOS, Linux, and Windows.

```{r}
# greta — requires Python + TensorFlow
# install.packages("greta")
# greta::install_greta_deps()  # can fail in many creative ways

# gretaR — pure R + torch
install.packages("gretaR")
# That's it. No Python, no conda, no reticulate.
```

### 2. Torch backend (not TensorFlow)

Under the hood, gretaR builds a computation graph using torch tensors. The
automatic differentiation engine is torch's autograd, which provides exact
gradients for HMC/NUTS without finite-difference approximations.

For most users this is invisible — the user-facing API is the same. But if you
wrote custom greta operations using `tensorflow` directly, those will need to
be rewritten using `torch`.

### 3. Additional inference methods

greta provides only `mcmc()` (HMC/NUTS). gretaR adds three more:

```{r}
m <- model(mu, sigma)

# MCMC (HMC or NUTS) — same as greta
draws <- mcmc(m, n_samples = 1000, warmup = 1000, chains = 4)

# Convenience wrappers
draws <- nuts(m, n_samples = 1000)
draws <- hmc(m, n_samples = 1000)

# MAP optimisation — find the posterior mode
fit <- opt(m)
fit$par

# Laplace approximation — Gaussian centred at MAP
la <- laplace(m)
la$mean
la$sd

# Variational inference (ADVI) — mean-field or full-rank Gaussian
vi_fit <- vi(m, method = "meanfield")
vi_fit$mean
vi_fit$draws  # samples from the variational posterior
```

### 4. Formula interface

gretaR includes `gretaR_glm()`, a high-level convenience function for fitting
Bayesian generalised linear models using standard R formula syntax:

```{r}
# Bayesian linear regression in one call
fit <- gretaR_glm(
  Sepal.Length ~ Sepal.Width + Petal.Length,
  data = iris,
  family = "gaussian"
)
summary(fit$draws)

# Logistic regression
fit <- gretaR_glm(y ~ x1 + x2, data = dat, family = "binomial")

# With custom priors
fit <- gretaR_glm(
  y ~ x,
  data = dat,
  family = "gaussian",
  prior = list(beta = normal(0, 1), sigma = half_cauchy(1))
)
```

There is no equivalent in greta — all models must be built from the DSL.

### 5. Renamed distribution constructors

greta uses names like `gamma()`, `beta()`, and `poisson()` that mask base R
functions. gretaR avoids this by appending `_dist`:

| greta | gretaR | Reason |
|:------|:-------|:-------|
| `gamma(shape, rate)` | `gamma_dist(shape, rate)` | Avoids masking `base::gamma()` |
| `beta(shape1, shape2)` | `beta_dist(alpha, beta)` | Avoids masking `base::beta()` |
| `poisson(lambda)` | `poisson_dist(rate)` | Avoids masking `stats::poisson()` |

### 6. Sparse matrix support

gretaR's `as_data()` accepts sparse matrices from the **Matrix** package
directly. This is important for large, sparse design matrices common in
ecology, genomics, and spatial models.

```{r}
library(Matrix)
X_sparse <- sparseMatrix(i = c(1, 2, 3), j = c(1, 2, 3), x = 1.0, dims = c(3, 3))
X <- as_data(X_sparse)
```

### 7. Posterior output format

greta returns `coda::mcmc.list` objects from `mcmc()`. gretaR returns
`posterior::draws_array` objects, which integrate directly with the modern
**posterior** and **bayesplot** ecosystems.

```{r}
# greta
draws <- greta::mcmc(m)
class(draws)
#> [1] "mcmc.list"
coda::summary(draws)

# gretaR
draws <- gretaR::mcmc(m)
class(draws)
#> [1] "gretaR_draws" "draws_array" "draws" "array"
posterior::summarise_draws(draws)
bayesplot::mcmc_trace(draws)
```

If your downstream code expects `coda::mcmc.list`, convert with:

```{r}
coda_draws <- posterior::as_mcmc.list(draws)
```

### 8. MCMC argument differences

The `mcmc()` signatures are similar but not identical:

| Argument | greta | gretaR | Default |
|:---------|:------|:-------|:--------|
| Samples | `n_samples` | `n_samples` | 1000 |
| Warmup | `warmup` | `warmup` | 1000 |
| Chains | `chains` | `chains` | 4 |
| Sampler | `sampler = hmc()` or `sampler = nuts()` | `sampler = "nuts"` or `"hmc"` | NUTS |
| Verbose | `verbose` | `verbose` | TRUE |
| Step size | Set inside `hmc()`/`nuts()` | `step_size` | Auto-tuned |
| Tree depth | Set inside `nuts()` | `max_treedepth` | 10 |
| Accept rate | Set inside sampler | `target_accept` | 0.8 (NUTS) |

In greta, sampler configuration is done via sampler objects passed to `mcmc()`.
In gretaR, all sampler options are direct arguments to `mcmc()`.

## Worked example: Bayesian linear regression

Below is a complete Bayesian linear regression in both greta and gretaR,
side by side.

### greta version

```{r}
library(greta)

# Simulate data
set.seed(42)
n <- 100
x_obs <- rnorm(n)
y_obs <- 2 + 1.5 * x_obs + rnorm(n, sd = 0.8)

# Wrap data
x <- as_data(x_obs)
y <- as_data(y_obs)

# Priors
alpha <- normal(0, 10)
beta  <- normal(0, 10)
sigma <- cauchy(0, 3, truncation = c(0, Inf))

# Linear predictor and likelihood
mu <- alpha + beta * x
distribution(y) <- normal(mu, sigma)

# Compile and sample
m <- model(alpha, beta, sigma)
draws <- mcmc(m, n_samples = 1000, warmup = 1000, chains = 4)

# Summarise
summary(draws)
```

### gretaR version

```{r}
library(gretaR)

# Simulate data
set.seed(42)
n <- 100
x_obs <- rnorm(n)
y_obs <- 2 + 1.5 * x_obs + rnorm(n, sd = 0.8)

# Wrap data
x <- as_data(x_obs)
y <- as_data(y_obs)

# Priors
alpha <- normal(0, 10)
beta  <- normal(0, 10)
sigma <- half_cauchy(3)     # <-- dedicated half-Cauchy constructor

# Linear predictor and likelihood
mu <- alpha + beta * x
distribution(y) <- normal(mu, sigma)

# Compile and sample
m <- model(alpha, beta, sigma)
draws <- mcmc(m, n_samples = 1000, warmup = 1000, chains = 4)

# Summarise — uses posterior package
posterior::summarise_draws(draws)

# Or use the formula interface for the same model in one call:
fit <- gretaR_glm(y_obs ~ x_obs,
                  data = data.frame(y_obs, x_obs),
                  family = "gaussian")
```

### What changed?

Only two lines differ:

1. `library(greta)` becomes `library(gretaR)`.
2. `cauchy(0, 3, truncation = c(0, Inf))` becomes `half_cauchy(3)`.

Everything else — `as_data()`, `normal()`, `variable()`, `distribution<-()`,
`model()`, `mcmc()` — works identically.

## Quick migration checklist

- [ ] Replace `library(greta)` with `library(gretaR)`
- [ ] Replace `gamma(...)` with `gamma_dist(...)`
- [ ] Replace `beta(...)` with `beta_dist(...)`
- [ ] Replace `poisson(...)` with `poisson_dist(...)`
- [ ] Replace truncated Cauchy/normal priors with `half_cauchy()` / `half_normal()`
- [ ] Replace `coda::` summary/plotting calls with `posterior::` / `bayesplot::` equivalents
- [ ] If using `greta::opt()` or `greta::calculate()`: check gretaR equivalents (`opt()`, `laplace()`, `vi()`)
- [ ] Remove `greta::install_greta_deps()` and any Python/conda setup from your workflow
- [ ] If using custom TensorFlow operations: rewrite using torch
