---
title: "gretaR: A Torch-Based Probabilistic Programming Framework for Bayesian Statistical Modelling in R"
subtitle: "Technical Documentation and Reference"
author:
  - name: Max Moldovan
    affiliation: Adelaide University, Australia
    orcid: 0000-0001-9680-8474
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: |
  gretaR is an R package for Bayesian statistical modelling that provides an
  R-native domain-specific language (DSL) for defining probabilistic models,
  compiled to differentiable computation graphs for inference via Hamiltonian
  Monte Carlo, the No-U-Turn Sampler, variational inference, and maximum a
  posteriori estimation. The package offers a dual-backend architecture: a
  native torch backend for GPU-capable custom models and an automatically
  generated Stan backend for production-speed inference on standard models.
  This document provides comprehensive technical documentation of the package
  architecture, algorithms, and implementation.
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    keep_tex: true
geometry: margin=2.5cm
fontsize: 11pt
bibliography: paper.bib
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, comment = "#>")
library(knitr)
```

# Introduction

gretaR is a probabilistic programming package for R that enables Bayesian
statistical modelling using native R syntax. Models are defined interactively
by manipulating `gretaR_array` objects --- data wrappers, random variables,
and deterministic transformations --- which lazily construct a directed
acyclic graph (DAG). This graph is then compiled into a differentiable
log-joint density function for gradient-based inference.

The package addresses three limitations of existing tools:

1. **greta** [@greta] provides an elegant R-native DSL but requires Python,
   TensorFlow, and reticulate --- a fragile dependency chain.
2. **Stan/brms** [@stan; @brms] requires learning a separate modelling language
   and managing a C++ toolchain.
3. **NIMBLE** [@nimble] uses BUGS-like syntax that differs from standard R.

gretaR resolves these by providing pure R syntax with a torch backend
(no Python), optional Stan backend (no new language), and five inference
methods in a single package.

## Scope of this Document

This document covers every major component of gretaR v0.1.0:

- The directed acyclic graph (DAG) and lazy evaluation system (Section 2)
- The 18 probability distributions and parameter transforms (Section 3)
- Five inference engines: NUTS, HMC, ADVI, MAP, Laplace (Section 4)
- Performance optimisations (Section 5)
- The Stan code generation backend (Section 6)
- The formula interface and random effects (Section 7)
- Custom distributions and mixture models (Section 8)
- Sparse matrix support (Section 9)
- The unified output structure (Section 10)
- Validation results (Section 11)

# Core Architecture: The Directed Acyclic Graph

## Lazy Evaluation and Node Types

gretaR uses lazy evaluation: operations on `gretaR_array` objects do not
compute values immediately. Instead, they construct a DAG of interconnected
nodes. Each node is an R6 object (`GretaRArray`) with fields:

- **`node_type`**: one of `"data"`, `"variable"`, or `"operation"`
- **`value`**: a torch tensor (fixed for data; managed by the sampler for variables)
- **`parents`**: character vector of parent node identifiers
- **`operation`**: a closure that computes this node's value from its parents
- **`distribution`**: a prior distribution (for variable nodes)
- **`constraint`**: bounds on the parameter space
- **`transform`**: a bijector mapping between constrained and unconstrained spaces

## Model Compilation

The `model()` function compiles the DAG into a differentiable log-joint
density function:

1. **Identify free variables**: traverse the DAG to find all variable nodes
2. **Exclude likelihood templates**: variable nodes used on the right-hand side
   of `distribution(y) <-` are not free parameters
3. **Build parameter layout**: compute offsets for packing/unpacking a flat
   parameter vector $\boldsymbol\theta_{\text{free}} \in \mathbb{R}^d$
4. **Construct `log_prob()`**: a function $f: \mathbb{R}^d \to \mathbb{R}$
   that computes:

$$\log p(\boldsymbol\theta, \mathbf{y}) = \underbrace{\sum_{k=1}^{K} \log p(\theta_k \mid \text{prior}_k)}_{\text{priors}} + \underbrace{\sum_{k=1}^{K} \log \left| \frac{\partial T_k^{-1}}{\partial \theta_k^{\text{free}}} \right|}_{\text{Jacobian}} + \underbrace{\sum_{i=1}^{N} \log p(y_i \mid \boldsymbol\theta)}_{\text{likelihood}}$$

where $T_k$ is the bijector for parameter $k$.

## Gradient Computation

Gradients are computed via torch's automatic differentiation:

```r
grad_log_prob <- function(model, theta_free) {
  theta_free$requires_grad_(TRUE)
  lp <- log_prob(model, theta_free)
  grads <- autograd_grad(lp, theta_free)
  list(lp = lp$item(), grad = grads[[1]])
}
```

The use of `autograd_grad()` (rather than `backward()`) avoids gradient
accumulation overhead and eliminates the need for gradient zeroing between
evaluations.

# Probability Distributions

gretaR implements 18 probability distributions, each as an R6 class
inheriting from `GretaRDistribution`. Every distribution provides:

- `log_prob(x)`: differentiable log-density evaluated at $x$ (returns a scalar torch tensor)
- `sample(n)`: draw $n$ samples (for prior predictive checks)
- `constraint`: parameter space bounds (used to select the appropriate bijector)

## Continuous Distributions

```{r dist-table, echo=FALSE, eval=TRUE}
dists <- data.frame(
  Distribution = c("Normal", "Half-Normal", "Half-Cauchy", "Student-$t$",
                    "Cauchy", "Exponential", "Gamma", "Beta",
                    "Log-Normal", "Uniform", "Multivariate Normal",
                    "Dirichlet", "LKJ Correlation", "Wishart"),
  Constructor = c("\\texttt{normal()}", "\\texttt{half\\_normal()}", "\\texttt{half\\_cauchy()}",
                   "\\texttt{student\\_t()}", "\\texttt{cauchy()}", "\\texttt{exponential()}",
                   "\\texttt{gamma\\_dist()}", "\\texttt{beta\\_dist()}",
                   "\\texttt{lognormal()}", "\\texttt{uniform()}",
                   "\\texttt{multivariate\\_normal()}", "\\texttt{dirichlet()}",
                   "\\texttt{lkj\\_correlation()}", "\\texttt{wishart()}"),
  Support = c("$\\mathbb{R}$", "$\\mathbb{R}^+$", "$\\mathbb{R}^+$",
               "$\\mathbb{R}$", "$\\mathbb{R}$", "$\\mathbb{R}^+$",
               "$\\mathbb{R}^+$", "$[0,1]$", "$\\mathbb{R}^+$",
               "$[a,b]$", "$\\mathbb{R}^k$", "Simplex",
               "Corr. matrix", "PD matrix"),
  Transform = c("Identity", "Log", "Log", "Identity", "Identity", "Log",
                 "Log", "Logit", "Log", "Scaled logit", "Identity",
                 "Identity*", "Identity*", "Identity*"),
  stringsAsFactors = FALSE
)
kable(dists, booktabs = TRUE, escape = FALSE,
      caption = "Continuous distributions. *Simplex, correlation, and PD transforms are deferred to a future release.")
```

## Discrete Distributions

```{r disc-table, echo=FALSE, eval=TRUE}
disc <- data.frame(
  Distribution = c("Bernoulli", "Binomial", "Poisson", "Negative Binomial"),
  Constructor = c("\\texttt{bernoulli()}", "\\texttt{binomial()}",
                   "\\texttt{poisson\\_dist()}", "\\texttt{negative\\_binomial()}"),
  Support = c("$\\{0,1\\}$", "$\\{0,\\ldots,n\\}$",
               "$\\mathbb{Z}_{\\geq 0}$", "$\\mathbb{Z}_{\\geq 0}$"),
  stringsAsFactors = FALSE
)
kable(disc, booktabs = TRUE, escape = FALSE,
      caption = "Discrete distributions (used as likelihoods; not directly sampled via HMC).")
```

## Parameter Transforms (Bijectors)

Six bijectors map between constrained parameter spaces and the unconstrained
real line, as required by gradient-based samplers:

| Bijector | Domain | Forward $T(x)$ | Inverse $T^{-1}(y)$ | $\log|J|$ |
|---|---|---|---|---|
| Identity | $\mathbb{R}$ | $y = x$ | $x = y$ | $0$ |
| Log | $\mathbb{R}^+$ | $y = \log x$ | $x = e^y$ | $\sum y_i$ |
| Logit | $(0,1)$ | $y = \log\frac{x}{1-x}$ | $x = \sigma(y)$ | $\sum[\log\sigma(y) + \log(1-\sigma(y))]$ |
| Scaled logit | $(a,b)$ | Shifted logit | Shifted sigmoid | Includes range adjustment |
| Softplus | $\mathbb{R}^+$ | $y = \log(e^x - 1)$ | $x = \text{softplus}(y)$ | $\sum\log\sigma(y)$ |
| Lower bound | $(L, \infty)$ | $y = \log(x - L)$ | $x = e^y + L$ | $\sum y_i$ |

The `select_transform()` function automatically chooses the bijector based on
the distribution's constraint specification.

# Inference Engines

## No-U-Turn Sampler (NUTS)

The primary inference method implements the NUTS algorithm
[@nuts] with the following components:

**Leapfrog integrator.** Given position $\boldsymbol\theta$ and momentum
$\mathbf{p}$, one step with step size $\epsilon$ and mass matrix $\mathbf{M}$:

\begin{align}
\mathbf{p} &\leftarrow \mathbf{p} + \frac{\epsilon}{2} \nabla_\theta \log p(\boldsymbol\theta, \mathbf{y}) \\
\boldsymbol\theta &\leftarrow \boldsymbol\theta + \epsilon \mathbf{M}^{-1} \mathbf{p} \\
\mathbf{p} &\leftarrow \mathbf{p} + \frac{\epsilon}{2} \nabla_\theta \log p(\boldsymbol\theta, \mathbf{y})
\end{align}

**Tree doubling.** The trajectory is extended by doubling in a randomly chosen
direction. At each depth $j$, $2^j$ leapfrog steps are evaluated. Growth
stops when a U-turn is detected:
$\Delta\boldsymbol\theta \cdot \mathbf{p}_{\text{left}} < 0$ or
$\Delta\boldsymbol\theta \cdot \mathbf{p}_{\text{right}} < 0$.

**Windowed warmup.** Adaptation follows a three-phase strategy inspired by
Stan [@stan]:

- Phase 1 (0--15\%): step-size adaptation via dual averaging [@nesterov2009]
- Phase 2 (15--90\%): collect samples for diagonal mass matrix estimation
- Phase 3 (90--100\%): re-adapt step size with the learned mass matrix

This phased approach prevents the mass matrix update from invalidating
the previously adapted step size.

**Implementation note.** The leapfrog operates on plain R numeric vectors
(not torch tensors) to avoid scoping issues with torch's `with_no_grad()`
in R. Only the gradient evaluation touches torch.

## Hamiltonian Monte Carlo (HMC)

Static HMC with a fixed number of leapfrog steps per iteration. Uses
identical warmup adaptation to NUTS. Recommended for debugging or when
NUTS tree depth is a concern.

## Automatic Differentiation Variational Inference (ADVI)

Implements the ADVI framework of @advi with two variational families:

**Mean-field:** $q(\boldsymbol\theta) = \mathcal{N}(\boldsymbol\mu, \text{diag}(\boldsymbol\sigma^2))$

**Full-rank:** $q(\boldsymbol\theta) = \mathcal{N}(\boldsymbol\mu, \mathbf{L}\mathbf{L}^\top)$

The ELBO is optimised via the reparameterisation trick:

$$\text{ELBO} = \mathbb{E}_{q}[\log p(\mathbf{y}, \boldsymbol\theta)] + \mathcal{H}[q]$$

where $\boldsymbol\theta = \boldsymbol\mu + \boldsymbol\sigma \odot \boldsymbol\epsilon$
and $\boldsymbol\epsilon \sim \mathcal{N}(\mathbf{0}, \mathbf{I})$.

Convergence is assessed by monitoring the relative change in the median ELBO
over a rolling window.

## MAP Estimation

Maximum a posteriori estimation via the Adam optimiser [@kingma2015adam].
The initial point is found by gradient ascent toward the posterior mode
(200 steps with learning rate 0.1), providing robust initialisation even
for multimodal posteriors.

## Laplace Approximation

Approximates the posterior as a multivariate normal centred at the MAP:

$$p(\boldsymbol\theta \mid \mathbf{y}) \approx \mathcal{N}\left(\hat{\boldsymbol\theta}_{\text{MAP}}, \left[-\nabla^2 \log p(\hat{\boldsymbol\theta}_{\text{MAP}}, \mathbf{y})\right]^{-1}\right)$$

The Hessian is computed via symmetric finite differences on the gradient.
Eigenvalue clamping ensures positive definiteness. The log marginal
likelihood is approximated as:

$$\log p(\mathbf{y}) \approx \log p(\hat{\boldsymbol\theta}, \mathbf{y}) + \frac{d}{2}\log(2\pi) + \frac{1}{2}\log|\boldsymbol\Sigma|$$

# Performance Optimisations

Three optimisations reduce the torch backend's per-gradient-evaluation cost:

## Compiled Log-Probability Function

At model compilation time, `compile_log_prob()` pre-extracts all parameter
layout information (offsets, dimensions, transforms, distribution references)
into a closure. This eliminates R6 method dispatch and list lookups during
the hot path of gradient evaluation.

## `autograd_grad()` vs `backward()`

The standard `backward()` accumulates gradients into the `$grad` attribute,
requiring subsequent `$clone()` and `$detach()$cpu()` calls. Using
`autograd_grad()` returns gradients directly, reducing per-call overhead
by approximately 27\%.

## JIT Tracing

The compiled log-probability function is optionally traced via
`torch::jit_trace()`, which fuses torch operations into an optimised graph.
Tracing succeeds for models without data-dependent control flow; a safe
fallback to the untraced compiled function is used otherwise.

**Combined effect:** 3--4$\times$ speedup over the initial implementation,
reducing the torch--Stan gap from 50--64$\times$ to 7--16$\times$ for
small models.

### Approaches Evaluated and Rejected

| Approach | Finding | Decision |
|---|---|---|
| GPU/MPS acceleration | 5.6$\times$ *slower* for small tensors due to transfer latency | Rejected |
| C++ leapfrog loop | R-level loop overhead is $<$1\% of total time | Rejected |
| In-place tensor operations | Higher dispatch cost than allocation for small tensors | Rejected |
| Batched chains (vectorised HMC) | NUTS trees vary per chain, wasting compute | Deferred |

# Stan Backend

## Motivation

The torch backend executes the DAG interpretively: each gradient evaluation
traverses R-level data structures and dispatches torch operations individually.
Stan compiles the entire model to optimised C++ with static automatic
differentiation, yielding 32--152$\times$ faster sampling on hierarchical
models.

gretaR's Stan backend provides this speed advantage while preserving the
R-native DSL:

```r
# Same model definition
mu <- normal(0, 10)
sigma <- half_cauchy(2)
distribution(y) <- normal(mu, sigma)
m <- model(mu, sigma)

# Choose backend at inference time
fit_torch <- mcmc(m, backend = "torch")  # native torch NUTS
fit_stan  <- mcmc(m, backend = "stan")   # compiled Stan NUTS
```

## Code Generation

`compile_to_stan()` walks the compiled model's DAG and emits valid Stan code:

1. **Data block**: declares observed data with dimensions inferred from tensor shapes
2. **Parameters block**: declares free parameters with constraints (e.g., `real<lower=0>`)
3. **Model block**: emits prior statements and likelihood statements

Each gretaR distribution maps to its Stan equivalent (e.g.,
`half_cauchy(scale)` $\to$ `cauchy(0, scale)` with `<lower=0>` bound).

## Benchmark Results

```{r bench-table, echo=FALSE, eval=TRUE}
bench <- data.frame(
  Model = c("Intercept-only (2p)", "Linear regression (3p)",
            "Random intercept LMM (23p)", "Crossed random effects (42p)",
            "Logistic GLMM (17p)", "Poisson GLMM (22p)",
            "Random slopes (35p)"),
  Torch = c(351.1, 2.9, 466.5, 1154.5, 423.1, 441.8, 208.8),
  Stan = c(3.7, 3.4, 4.6, 7.6, 8.6, 5.9, 6.5),
  stringsAsFactors = FALSE
)
bench$Speedup <- paste0(round(bench$Torch / bench$Stan, 0), "x")
names(bench) <- c("Model", "Torch (s)", "Stan (s)", "Speedup")
kable(bench, booktabs = TRUE,
      caption = "Wall-clock time (500 warmup + 500 samples, 2 chains). Parameter recovery matches to 3 significant figures across backends.")
```

# Formula Interface

## Basic Usage

`gretaR_glm()` provides a high-level interface using standard R formula syntax:

```r
fit <- gretaR_glm(
  Sepal.Length ~ Sepal.Width + Petal.Length,
  data = iris,
  family = "gaussian"
)
```

Internally, `model.matrix()` constructs the design matrix; default priors
are assigned ($\beta \sim \mathcal{N}(0, 5)$, $\sigma \sim \text{Half-Cauchy}(2)$);
and the model is compiled and sampled via the standard gretaR pipeline.

## Random Effects

lme4-style random effects are parsed via regex (no lme4 dependency required):

```r
fit <- gretaR_glm(
  y ~ x + (1 | group),        # random intercepts
  data = dat,
  family = "gaussian"
)
```

Supported patterns:

- `(1 | group)` --- random intercepts
- `(x | group)` --- correlated random intercepts and slopes
- `(0 + x | group)` --- random slopes only
- Multiple terms: `(1 | site) + (1 | year)`

All random effects use non-centred parameterisation by default:
$\alpha_j = \tau \cdot z_j$ where $z_j \sim \mathcal{N}(0, 1)$ and
$\tau \sim \text{Half-Cauchy}(2)$. This avoids the funnel geometry
that degrades HMC performance in centred parameterisations [@betancourt2015].

## Formula Style Detection

The parser auto-detects formula styles:

| Pattern | Detected as | Action |
|---|---|---|
| `y ~ x1 + x2` | base R | Standard GLM via `model.matrix()` |
| `y ~ x + (1\|group)` | lme4 | Mixed model with non-centred RE |
| `y ~ s(x)` | mgcv | Error (not yet supported) |

# Custom Distributions and Mixture Models

## Custom Distributions

Users can define distributions with arbitrary torch-differentiable
log-probability functions:

```r
x <- custom_distribution(
  log_prob_fn = function(x) -torch_sum(torch_abs(x)),
  constraint = list(lower = -Inf, upper = Inf),
  name = "laplace"
)
```

The supplied function must accept a torch tensor and return a scalar
torch tensor. It must be differentiable via torch autograd.

## Mixture Models

Finite mixture models are supported via the log-sum-exp trick:

$$\log p(x \mid \boldsymbol\pi, \boldsymbol\Theta) = \log \sum_{k=1}^{K} \pi_k \, p_k(x \mid \theta_k) = \text{logsumexp}_k \left[\log \pi_k + \log p_k(x \mid \theta_k)\right]$$

This marginalises over the discrete component indicator, enabling
gradient-based inference:

```r
w <- dirichlet(c(1, 1))
mu1 <- normal(-2, 1); mu2 <- normal(2, 1)
sigma <- half_cauchy(1)

mix <- mixture(
  distributions = list(normal(mu1, sigma), normal(mu2, sigma)),
  weights = w
)
distribution(y) <- mix
```

# Sparse Matrix Support

Large, sparse design matrices (common in genomics, NLP, and spatial modelling)
are handled via the Matrix package [@matrix]. The `as_data()` function
dispatches on `sparseMatrix` objects:

```r
library(Matrix)
X_sparse <- sparseMatrix(i = ..., j = ..., x = ...)
X <- as_data(X_sparse)  # stored as torch sparse COO tensor
mu <- X %*% beta         # sparse-aware matrix multiplication
```

The Matrix package is a recommended R package shipped with every R installation,
imposing zero additional installation burden. Sparse $\times$ dense matrix
multiplication uses `torch_mm()`, which handles the sparse layout natively.

# Unified Output Structure

All inference functions return a `gretaR_fit` S3 object with consistent
structure:

```r
fit$draws        # posterior::draws_array (NULL for MAP/Laplace)
fit$model        # compiled gretaR_model
fit$summary      # tibble: mean, sd, quantiles, R-hat, ESS
fit$convergence  # list: n_eff, rhat, max_rhat, min_ess, n_divergences
fit$call_info    # original arguments for reproducibility
fit$run_time     # elapsed seconds
fit$method       # "nuts", "hmc", "vi", "map", "laplace", "stan"
fit$par          # point estimates (all methods)
```

S3 methods provide a consistent interface:

- `print(fit)` --- concise summary with convergence diagnostics
- `summary(fit)` --- full posterior table via `posterior::summarise_draws()`
- `coef(fit)` --- named vector of posterior means or MAP estimates
- `plot(fit, type)` --- diagnostic plots via bayesplot (`"trace"`, `"density"`,
  `"pairs"`, `"rhat"`, `"neff"`)

# Validation

gretaR was validated against CmdStan [@stan] on 10 benchmark models.
Parameter estimates agree to 2--3 significant figures across all benchmarks.
The validation suite is included in `inst/validation/` and covers:

- Normal mean estimation (known and unknown variance)
- Linear and multiple regression
- Logistic and Poisson regression
- Gamma GLM and robust regression (Student-$t$ errors)
- Beta regression
- Hierarchical random intercepts (non-centred parameterisation)

## Correctness Criteria

A model is considered validated when:

- $\hat{R} < 1.05$ for all parameters
- Bulk ESS $> 400$ for all parameters
- Posterior means within MCMC uncertainty of Stan reference values
- Zero post-warmup divergent transitions

# Potential for Further Development

## Near-Term (Phase 3)

- **mgcv smooth terms** (`s()`, `te()`) via basis function expansion
- **Simplex and correlation transforms** for Dirichlet and LKJ sampling
- **Wishart/inverse-Wishart sampling** via Bartlett decomposition
- **Improved Stan code generator** with operation-type metadata on DAG nodes

## Medium-Term

- **Gaussian Process module** (`gretaR.gp`) as an extension package
- **ODE-based models** using torch ODE solvers
- **Expectation propagation** for approximate inference
- **S7 class system** migration (when mature)

## Long-Term

- **GPU-accelerated inference** for large latent variable models
- **Batched chain evaluation** for static HMC on GPU
- **Community extension framework** with templates and documentation
- **CRAN submission** and JOSS publication

# References
